Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-01-28 3.phase_transitions/3.global-tests.tex
updated Ch 4
I updated up to and including Ch 4.2.2.

Hope to finish tomorrow.
Chaoter 3
Fixed some mathematical typos and incorporated the proof of Theorem 3.1 in the text.

Also "clarified" the statement of Theorem 3.1.  Please check.

\begin{theorem} \label{thm:detection-optimality}
Consider the signal detection problem in the triangular array of Gaussian error models \eqref{eq:model-additive-Chapter3} where the sparsity is parametrized as in \eqref{eq:signal-sparsity-additive}.
\begin{itemize}
    \item For signals whose sizes are parametrized as in \eqref{eq:signal-size-additive}, the detection problem can be asymptotically solved in the sense of \eqref{eq:support-recovery-success} with $L_2$, $L_1$, or $S$ statistic when $\beta\le 1/2$; on the other hand, these statistics are asymptotically powerless in the sense of \eqref{eq:support-recovery-failure} when $\beta>1/2$.
    \item For small and dense signals whose signal sizes are parametrized as in \eqref{eq:signal-size-small}, the detection problem can be asymptotically solved in the sense of \eqref{eq:support-recovery-success} with $L_2$ or $L_1$ statistic when $\underline{r}>\beta/2-{1}/{4}$; on the other hand, these statistics are asymptotically powerless in the sense of \eqref{eq:support-recovery-failure} when $\overline{r}<\beta/2-{1}/{4}$.
    Further, tests based on the $S$ statistic can succeed asymptotically in the sense of \eqref{eq:support-recovery-success} when $\underline{r}>\beta-1/2$, hence attaining the boundary of detectability in \eqref{eq:detection-boundary-small-signals}.
\end{itemize}
\end{theorem}

Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:supplement}.
We visualize the results in Theorem in Figure \ref{fig:phase-diagram-signal-detection}.
It is worth noting that the $\beta$-$r$ parameter regions where $L_1$ and $L_2$ statistics are asymptotically powerful coincide, and these statistics are theoretically suboptimal for both sparse regimes ($\beta>1/2$) and relatively dense regimes ($\beta\le 1/2$).

@@

Optimality properties of such adaptive tests and the optimal choice of the $q$-combinations, however, remain open problems.
\cite{xu2016adaptive} suggested combine $q=1, 2, 3, \ldots, 6$, and  $q=\infty$, based empirical evidence from numerical experiments. 
Theorem \ref{thm:detection-optimality} here implies that, at least for detecting one-sided alternatives, the $\widetilde{L}_2$ statistic (i.e., $L_2$ norm) and the $L_1$ norm are asymptotically dominated by the $\widetilde{L}_1$ statistic (or equivalently, the sum $S$). Therefore it is sufficient to include only the latter in the construction of the adaptive test.

>>>>

\begin{theorem} \label{thm:detection-optimality}
Consider the signal detection problem in the triangular array of Gaussian error models \eqref{eq:model-additive-Chapter3} where the sparsity is parametrized as in \eqref{eq:signal-sparsity-additive}.\\

{\em (i)}  For $\beta \in (1/2,1)$ and growing signals sizes as in \eqref{eq:signal-size-additive}, the statistics
    $L_1, L_2$ and $T$ are asymptotically powerless in the sense of \eqref{eq:support-recovery-failure}.\\

{\em (ii)} For $\beta \in (0,1/2]$ and growing signals sizes as in \eqref{eq:signal-size-additive}, the statistics
    $L_1, L_2$ and $T$ solve the detection problem in the sense of \eqref{eq:support-recovery-success}.\\

{\em (iii)} For dense and faint signals, i.e., $\beta \in (0,1/2]$ under the parameterization \eqref{eq:signal-size-small}, the sum statistic $T$
    attains the optimal detectability boundary in \eqref{eq:detection-boundary-small-signals}.  That is, tests based on the sum statistic $T$ can 
    succeed asymptotically in the sense of \eqref{eq:support-recovery-success} when $\underline{r}>\beta-1/2$.\\

 {\em (iv)} In the dense and faint signal setting of {\em (iii)}, the $L_1$ and $L_2$ statistics are both 
 sub-optimal. More precisely, they succeed in the sense of  \eqref{eq:support-recovery-success},  for $\underline{r}>\beta/2-{1}/{4}$, but fail in the 
 sense of \eqref{eq:support-recovery-failure} when $\overline{r}<\beta/2-{1}/{4}$.
\end{theorem}
%Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:supplement}.

\begin{proof} The claims in parts {\rm (i)} and {\rm (ii)} about the $L_1$, $L_2$, and the sum statistic $T$ in the cases of diverging signal sizes 
\eqref{eq:signal-size-additive} can be found in \cite{fan1996test} and \cite{candes2018lecture}. We prove here the statements for the cases {\rm (iii)}
and {\rm (iv)}, where the signals are dense and small, as parametrized in \eqref{eq:signal-size-small}.

\stilian{For simplicity of the exposition, we will suppose that  in \eqref{eq:signal-size-small} we have 
$\underline{r} = r = \overline{r}$, so that  $\mu(i) = p^r$.  The general case where $\underline{r}<\overline{r}$ is left as 
an exercise.}{\fbox{How about that?}}

{\em Part {\rm (iii)}:} We first show that the sum statistic $T$, or equivalently, the simple arithmetic mean attains the sparse signal 
detection boundary.  By the normality and independence of the summands, we have
\begin{equation}
    \frac{1}{\sqrt{p}}\sum_{i=1}^p x(i) \sim 
    \begin{cases}
    \mathrm{N}(0, 1), & \text{under } H_0\\
    \mathrm{N}(p^{(r-\beta)+1/2}, 1), & \text{under }  H_1.
    \end{cases}
\end{equation}
It immediately follows that the two distributions can be distinguished perfectly if $p^{r-(\beta-1/2)}$ diverges, i.e., $r>\beta-1/2$. 
This can be seen by simply setting the rejection region at $(p^{(r-\beta)+1/2}/2,\,+\infty)$ for the scaled statistic $\sum_{i=1}^p x(i)/\sqrt{p} $.
According to the lower bound on the performance limit in detection problems \citep[see Theorem 8 in][]{cai2011optimal}, we have shown that $T$ attains the optimal detection boundary \eqref{eq:detection-boundary-small-signals}.

{\em Part {\rm (iv)}:}  We now turn to the $L_2$-norm statistic.
Recall a non-central chi-square random variable $\chi^2_k(\lambda)$ has mean $(k+\lambda)$ and variance $2(k+2\lambda)$.
Since the observations have distributions $\mathrm{N}(0,1)$ under the null and $\mathrm{N}(p^r,1)$ under the alternative, we have $x^2(i)\sim \chi^2_1(0)$ for $i\not\in S$ and $x^2(i)\sim \chi^2_1(p^{2r})$ for $i\in S$.
Therefore, the mean and variance of the (centered and scaled) $L_2$ statistics are
\begin{equation}
    \E\left[\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right] = 
    \begin{cases}
    0 & \text{under } H_0\\
    p^{1-\beta}p^{2r}p^{-1/2} = p^{1/2-\beta+2r} & \text{under } H_1,
    \end{cases}
\end{equation}
and 
\begin{equation}\label{e:detection-variance}
    \mathrm{Var}\left(\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right) = 
    \begin{cases}
    \frac{1}{p}2p = 2 & \text{under } H_0\\
    \frac{1}{p}\left(2p+4p^{1-\beta+2r}\right) = 2(1+ 2p^{2r-\beta}) & \text{under } H_1,
    \end{cases}
\end{equation}
respectively.
By the central limit theorem, we have
\begin{equation}
    \frac{1}{\sqrt{2p}}\sum_{i=1}^p \left(x(i)^2-1\right) \implies \mathrm{N}(0,1),  
\end{equation}
under the null.    On the other hand, under the alternative, since $p^{2r-\beta}\to0$ for all $r<0$ and $\beta>0$, 
 the variance in \eqref{e:detection-variance} converges to $2$, as $p\to\infty$ and an application of the Lyapunov version of 
 the CLT, entails
 \begin{equation}
    \frac{1}{\sqrt{2p}}\left(\sum_{i=1}^p \left(x(i)^2-1\right) - p^{1/2-\beta+2r}\right) \implies \mathrm{N}(0,1).
\end{equation}
Hence, perfect detection with the $L_2$-norm is possible if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, if $r<\beta/2-1/4$, the distributions of the (scaled) statistics merge under the null and the alternative. % \fbox{Hellinger distance is scale-invariant (show), detection is impossible}.

The case of the $L_1$-norm is treated similarly.
Let $Y=|X|$ where $X\sim|\mathrm{N}(\mu,1)|$. 
Using the expressions for the mean and variance of $Y$ \citep[see, e.g.,][]{tsagris2014folded},
\begin{align}
    \mu_Y &= \E[Y] = \sqrt{\frac{2}{\pi}}e^{-\mu^2/2} + \mu(1-\Phi(-\mu)), \\
    \sigma_Y^2 &= \mathrm{Var}(Y) = \mu^2 + 1 - \mu^2_Y,
\end{align}
where $\Phi$ is the CDF of a standard normal random variable,
we have, regardless of the value of $\mu$,
\begin{equation} \label{eq:bounded-variance-abs-Gaussian}
    \sigma_Y^2 = \mathrm{Var}(Y) = \E(Y-\E Y)^2 \le \E(X-\E X)^2 = 1,
\end{equation}
where the inequality holds because absolute value is a Lipschitz function with Lipschitz constant 1.

By the central limit theorem, we have,
\begin{equation}
    \frac{1}{\sqrt{p}}\left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \implies \mathrm{N}(0, 1-2/\pi)
\end{equation}
under the null.
On the other hand, when the alternative hypothesis holds, we have
\begin{align*}
    \E \left[ \frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right] 
    &= \frac{p^{1-\beta}}{\sqrt{p}}\left[\left(\sqrt{\frac{2}{\pi}}e^{-\mu^2/2} + \mu\left(1-2\Phi(-\mu)\right) \right) - \sqrt{\frac{2}{\pi}} \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(e^{-p^{2r}/2} - 1 \right) + p^r\left(1-2\Phi(-\mu)\right) \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(-p^{2r}/2 - O(p^{4r})\right) + p^r\left(\sqrt{\frac{2}{\pi}}p^r + O(p^{3r})\right) \right] \\
    &= p^{1/2-\beta} \sqrt{\frac{2}{\pi}} \left(p^{2r}/2 + O(p^{4r}) \right) \\
    &= p^{1/2-\beta+2r} \sqrt{{1}/{2\pi}} + O(p^{1/2-\beta+4r}),
\end{align*}
and 
\begin{align*}
    \mathrm{Var} \left(\frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right) 
    & = {\color{red}\frac{1}{p}}  (p-p^{1-\beta})(1-2/\pi) + {\color{red}\frac{1}{p}}   p^{1-\beta}\sigma_{Y}^2 \\
    & \to 1-2/\pi,
\end{align*}
by the boundedness of $\sigma_{Y}^2$ shown in \eqref{eq:bounded-variance-abs-Gaussian}.
Again, by the Lyapunov version of the central limit theorem, we conclude asymptotic normality of the centered and scaled $L_1$-norms under the alternative.
In an entirely analogous argument to the $L_2$-norm case, asymptotically perfect detection can be achieved if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, when $r<\beta/2-1/4$, the two hypotheses cannot be told apart by the $L_1$-norms since the distributions of the (scaled) statistics merge under the two hypotheses.
\end{proof}

The portmanteau of results in Theorem \ref{thm:detection-optimality} are visualized in Figure \ref{fig:phase-diagram-signal-detection}.
It is worth noting that the $\beta$-$r$ parameter regions where $L_1$ and $L_2$ statistics are asymptotically powerful coincide, and 
these statistics are theoretically suboptimal for both sparse regimes ($\beta>1/2$) and relatively dense regimes ($\beta\le 1/2$).

@@

Optimality properties of such adaptive tests and the optimal choice of the $q$-combinations, however, remain open problems.
\cite{xu2016adaptive} suggested combining $q=1, 2, 3, \ldots, 6$, and  $q=\infty$, based empirical evidence from numerical experiments. 
Theorem \ref{thm:detection-optimality} here implies that, at least for detecting one-sided alternatives, the $\widetilde{L}_2$ statistic (i.e., $L_2$ norm) and the $L_1$ norm are asymptotically dominated by the $\widetilde{L}_1$ statistic (or equivalently, the sum $T$). Therefore it is sufficient to include only the latter in the construction of the adaptive test.

### 2021-01-28 5.Bayes.optimality/5.Bayes_optimality_and_minimax.tex
Chaoter 3
Fixed some mathematical typos and incorporated the proof of Theorem 3.1 in the text.

Also "clarified" the statement of Theorem 3.1.  Please check.

To this end, we begin by characterizing the finite-sample Bayes optimality of the thresholding procedures (Section
\ref{subsec:optimal-procedure-sub-exponential}). As we will see, the so-called oracle thresholding estimators are in 
fact finite-sample optimal for certain classes of models.  These optimality results allow us to establish 
powerful minimax formulations of the exact support recovery phase-transition phenomenon showing its universality for all estimators (Theorem \ref{thm:necessary-strengthened}).

Perhaps surprisingly, however, thresholding estimators can be sub-optimal.  This is so for example 
in the additive noise model with heavier than exponential error densities (Example 
\ref{exmp:suboptimal-data-thresholding}).   In this case, the likelihood ratio
thresholding rather than data thresholding can lead to optimal support estimators
(Section \ref{subsec:optimal-procedure-super-exponential}).  In such cases, the universality 
of the exact support recovery boundary remains an open question.

@@

Since $x(j) < x(j')$, and since each of $\{f_0, f_{i}\}$ is an MLR family, we have
$$
\frac{f_i(x(j))}{f_0(x(j))} - \frac{f_i(x(j'))}{f_0(x(j'))} \le 0 \implies f_i(x(j)) f_0(x(j')) - f_i(x(j')) f_0(x(j)) \le 0.
$$
Using Relation \eqref{eq:MLR-optimality-proof}, we conclude that $f(x|\widehat{S}) \le f(x|\widehat{S}\,')$.
\stilian{Continuing this way, we can successively improve the likelihood of every estimator until we arrive at an oracle thresholding estimator, proving the desired optimality.  Note that with the same argument, any two oracle 
thresholding estimators have the same likelihood.}{
\fbox{replacing} This implies that any estimator that is not $\widehat{S}^*$ may be improved, and the optimality follows.}
\end{proof}

@@

\begin{remark}[An open problem] As we have seen, the thresholding estimators are no longer optimal 
in the additive model with error-densities heavier than exponential.   Thanks to Proposition 
\ref{prop:likelihood-ratio-thresholding}, the oracle likelihood thresholding procedures are promising alternatives
that can lead us to practical support estimators.

In the case where the signals have difference sizes, however, the argument in the proof of Proposition \ref{prop:likelihood-ratio-thresholding} breaks downs since the signal densities need to be identical for Relation \eqref{eq:likelihood-ratio-thresholding-proof} to hold.  In such cases, the characterization of the optimal procedure is an open problem.
\end{remark}

>>>>

To this end, we begin by characterizing the finite-sample Bayes optimality of the thresholding procedures (Section
\ref{subsec:optimal-procedure-sub-exponential}). As we will see, the so-called oracle thresholding estimators are in 
fact finite-sample optimal for many classes of models.  These optimality results allow us to establish 
powerful minimax formulations of the exact support recovery phase-transition phenomenon showing its universality for all estimators (Theorem \ref{thm:necessary-strengthened}).

Perhaps surprisingly, however, thresholding estimators can be sub-optimal.  This is so for example 
in the additive noise model with heavier than exponential error densities (Example 
\ref{exmp:suboptimal-data-thresholding}).   In this case, {\em likelihood ratio
thresholding} rather than {\em data thresholding} can lead to optimal support estimators
(Section \ref{subsec:optimal-procedure-super-exponential}).  In such cases, the universality 
of the exact support recovery boundary remains an open question.

@@

Since $x(j) < x(j')$, and since each of $\{f_0, f_{i}\}$ is an MLR family, we have
$$
\frac{f_i(x(j))}{f_0(x(j))} - \frac{f_i(x(j'))}{f_0(x(j'))} \le 0 \implies f_i(x(j)) f_0(x(j')) - f_i(x(j')) f_0(x(j)) \le 0.
$$
Using Relation \eqref{eq:MLR-optimality-proof}, we conclude that $f(x|\widehat{S}) \le f(x|\widehat{S}\,')$.
\stilian{Continuing this way, we can successively improve the likelihood of every estimator until we arrive at an oracle thresholding estimator, proving the desired optimality.  Note that with the same argument, we obtain that any two oracle 
thresholding estimators have the same likelihood.}{
\fbox{replacing} This implies that any estimator that is not $\widehat{S}^*$ may be improved, and the optimality follows.}
\end{proof}

@@

\begin{remark}[{\bf An open problem\!}\,] As we have seen, the thresholding estimators are no longer optimal 
in the additive model with error-densities heavier than exponential.   Thanks to Proposition 
\ref{prop:likelihood-ratio-thresholding}, the oracle likelihood thresholding procedures are promising alternatives
that can lead us to practical support estimators.

In the case where the signals have different sizes, however, the argument in the proof of Proposition \ref{prop:likelihood-ratio-thresholding} breaks downs since the signal densities need to be identical for Relation \eqref{eq:likelihood-ratio-thresholding-proof} to hold.  In such cases, the characterization of the optimal procedure is an open problem.
\end{remark}
